National Longitudinal Evidence for Growth in Subjective Well-being from Spiritual Beliefs

### 2021-03-11 Manuscript_Bayesian.Rmd
Updates from Overleaf

keywords          : "Belief, God, Health, Longitudinal, Panel, Religion, Spirit, Spirituality, Well-being"
wordcount         : "5161 words in text body, 1191 words in reference section"
bibliography      : ["b-references.bib"]
floatsintext      : no
figurelist        : no
figsintext        : yes
tablelist         : no
footnotelist      : no
linenumbers       : no
mask              : no
draft             : no
documentclass     : "apa6"
classoption       : "man"
output            : papaja::apa6_pdf
extra_dependencies: ["rotating", "booktabs", "graphicx","longtable","amsmath","amsfonts","amssymb"]

---
```{r libraries}
library("tidyverse")
library("papaja")
library("here")

```
```{r global_options, echo=FALSE, include = FALSE}
knitr::opts_chunk$set(#fig.path="Figs/", 
                      message=FALSE, 
                      warning=FALSE,
                      collapse =TRUE,
                      echo=TRUE, 
                      #results="hide", 
                     # fig.width= 10,
                     # fig.height=8,
                      tidy = "styler")
```

@@

Here, we addresses these three challenges of previous research in the following ways: First, to avoid a tautological elision of "spirituality" and "well-being" we adopt a non-affective measure of spirituality as a "believe in a Spirit or Life Force." On the other side, to measure subjective well-being, we use a well-validated scale for personal well-being and a well-validated scale for Life-Satisfaction (see Method). Second, to disentangle spiritual beliefs from traditional religious beliefs and dis0beliefs we adopt a four-level categorical indicator combining the binary indicator of belief in a Spirit or Life Force with the binary indicator "Do you believe in some form of a spirit or Life Force?" The levels of this four-level categorical indicator for spiritual and religious beliefs and dis-beliefs are as follows: (1) **God and Spirit Disbelievers**: those who neither believe in a God nor believe in a Spirit or Life Force; (2) **Spirit Believers/God Disbelievers**: those who disbelieve in a God and believe in a Spirit or Life Force; (3) **God Believers/Spirit Disbelievers**: those who believe in a God but do not believe in a Spirit or Life Force; (4) **God and Spirit Believers**: Those who believe in a both God and a Spirit or Life Force. Clearly such an approach to categorising religion and spirituality cannot simultaneously address many interests in previous spirituality/well-being research. For example we do not combine our cognitive measure of belief with behavioral measures or attitudinal measures. However, this cost to scope is paid for with benefits to precision. The spiritual indicators measure cognitive states in which distinct varieties of religious and spiritual beliefs and dis-beliefs are straightforwardly disentangled and measured within people, and their longitudinal associations with reliable well-being indicators are systematically quantified Third, to understand how spirit beliefs affect growth, stability, or decline in psychological well-being over time, we leverage longitudinal responses from a nationally diverse probability sample year over nine years (2010 to 2018 New Zealand Attitudes and Values Study, NZAVS).

@@

\begin{table}[h]
\begin{center}
\scalebox{0.45}{
\begin{tabular}[t]{llllllllllllllllllll}
\toprule
\multicolumn{2}{c}{ } & \multicolumn{2}{c}{2010 (N=3820)} & \multicolumn{2}{c}{2011 (N=6153)} & \multicolumn{2}{c}{2012 (N=10385)} & \multicolumn{2}{c}{2013 (N=15836)} & \multicolumn{2}{c}{2014 (N=15026)} & \multicolumn{2}{c}{2015 (N=13665)} & \multicolumn{2}{c}{2016 (N=18071)} & \multicolumn{2}{c}{2017 (N=15779)} & \multicolumn{2}{c}{2018 (N=16501)} \\
\cmidrule(l{3pt}r{3pt}){3-4} \cmidrule(l{3pt}r{3pt}){5-6} \cmidrule(l{3pt}r{3pt}){7-8} \cmidrule(l{3pt}r{3pt}){9-10} \cmidrule(l{3pt}r{3pt}){11-12} \cmidrule(l{3pt}r{3pt}){13-14} \cmidrule(l{3pt}r{3pt}){15-16} \cmidrule(l{3pt}r{3pt}){17-18} \cmidrule(l{3pt}r{3pt}){19-20}
  &    & Mean & Std. Dev. & Mean  & Std. Dev.  & Mean   & Std. Dev.   & Mean    & Std. Dev.    & Mean     & Std. Dev.     & Mean      & Std. Dev.      & Mean       & Std. Dev.       & Mean        & Std. Dev.        & Mean         & Std. Dev.        \\
\midrule
Age &  & 51.0 & 14.7 & 50.9 & 15.5 & 50.2 & 14.8 & 48.6 & 14.0 & 49.7 & 14.0 & 51.0 & 13.9 & 50.9 & 13.6 & 51.9 & 13.6 & 52.7 & 13.5\\
NZdep &  & 4.8 & 2.8 & 4.6 & 2.7 & 4.9 & 2.8 & 4.7 & 2.8 & 4.7 & 2.8 & 4.7 & 2.8 & 4.6 & 2.7 & 4.6 & 2.7 & 4.6 & 2.7\\
Education &  &  &  &  &  & 5.8 & 2.8 & 6.0 & 2.8 & 6.1 & 2.8 & 6.2 & 2.8 & 6.4 & 2.7 & 6.5 & 2.7 & 6.5 & 2.7\\
Political\_Conservativism &  & 3.9 & 1.3 & 3.7 & 1.4 & 3.6 & 1.3 & 3.6 & 1.3 & 3.6 & 1.3 & 3.6 & 1.3 & 3.6 & 1.4 & 3.6 & 1.4 & 3.6 & 1.4\\
LIFESAT &  & 5.2 & 1.2 & 5.1 & 1.2 & 5.1 & 1.2 & 5.2 & 1.2 & 5.2 & 1.2 & 5.2 & 1.2 & 5.2 & 1.2 & 5.3 & 1.2 & 5.3 & 1.2\\
PWI &  & 6.8 & 1.7 & 6.8 & 1.7 & 6.8 & 1.7 & 7.0 & 1.7 & 7.0 & 1.7 & 7.1 & 1.7 & 7.1 & 1.7 & 7.1 & 1.7 & 7.1 & 1.6\\
\midrule
 &  & N & \% & N & \% & N & \% & N & \% & N & \% & N & \% & N & \% & N & \% & N & \%\\
Employed & 0 & 793 & 0.7 & 1276 & 1.1 & 2666 & 2.3 & 3531 & 3.1 & 3262 & 2.8 & 2972 & 2.6 & 4035 & 3.5 & 3655 & 3.2 & 3746 & 3.3\\
 & 1 & 2977 & 2.6 & 4795 & 4.2 & 7566 & 6.6 & 11991 & 10.4 & 11626 & 10.1 & 10378 & 9.0 & 14016 & 12.2 & 12070 & 10.5 & 12250 & 10.6\\
Ethnic\_Categories & Euro & 2947 & 2.6 & 4375 & 3.8 & 7881 & 6.8 & 12270 & 10.6 & 12051 & 10.5 & 10959 & 9.5 & 14641 & 12.7 & 12906 & 11.2 & 13475 & 11.7\\
 & Maori & 556 & 0.5 & 638 & 0.6 & 1489 & 1.3 & 1897 & 1.6 & 1830 & 1.6 & 1608 & 1.4 & 1964 & 1.7 & 1826 & 1.6 & 1872 & 1.6\\
 & Pacific & 110 & 0.1 & 126 & 0.1 & 343 & 0.3 & 408 & 0.4 & 400 & 0.3 & 336 & 0.3 & 385 & 0.3 & 286 & 0.2 & 296 & 0.3\\
 & Asian & 124 & 0.1 & 190 & 0.2 & 409 & 0.4 & 579 & 0.5 & 572 & 0.5 & 473 & 0.4 & 664 & 0.6 & 574 & 0.5 & 579 & 0.5\\
Male & 0 & 2390 & 2.1 & 3862 & 3.4 & 6542 & 5.7 & 10033 & 8.7 & 9500 & 8.2 & 8548 & 7.4 & 11377 & 9.9 & 9949 & 8.6 & 10388 & 9.0\\
 & 1 & 1430 & 1.2 & 2291 & 2.0 & 3843 & 3.3 & 5803 & 5.0 & 5478 & 4.8 & 5072 & 4.4 & 6635 & 5.8 & 5776 & 5.0 & 6077 & 5.3\\
Has\_Partner & 0 & 1054 & 0.9 & 1806 & 1.6 & 2988 & 2.6 & 4237 & 3.7 & 3664 & 3.2 & 3407 & 3.0 & 4274 & 3.7 & 3775 & 3.3 & 3931 & 3.4\\
 & 1 & 2764 & 2.4 & 4325 & 3.8 & 7331 & 6.4 & 11493 & 10.0 & 11130 & 9.7 & 10176 & 8.8 & 13613 & 11.8 & 11964 & 10.4 & 12078 & 10.5\\
Urban & Not\_Urban & 1531 & 1.3 & 2010 & 1.7 & 3512 & 3.0 & 5116 & 4.4 & 4894 & 4.2 & 4541 & 3.9 & 6223 & 5.4 & 2873 & 2.5 & 2938 & 2.5\\
 & Urban & 2271 & 2.0 & 3695 & 3.2 & 6709 & 5.8 & 10602 & 9.2 & 9937 & 8.6 & 8815 & 7.6 & 11622 & 10.1 & 12687 & 11.0 & 13291 & 11.5\\
Beliefs & \_Disbeliefs\_ & 625 & 0.5 & 1312 & 1.1 & 2235 & 1.9 & 3639 & 3.2 & 0 & 0.0 & 3278 & 2.8 & 4420 & 3.8 & 4135 & 3.6 & 4514 & 3.9\\
 & \_SpiritExcludesGod\_ & 778 & 0.7 & 1508 & 1.3 & 2535 & 2.2 & 4206 & 3.6 & 0 & 0.0 & 3542 & 3.1 & 4808 & 4.2 & 4148 & 3.6 & 4309 & 3.7\\
 & GodAndSpirit & 1751 & 1.5 & 2560 & 2.2 & 4181 & 3.6 & 6201 & 5.4 & 0 & 0.0 & 5780 & 5.0 & 7002 & 6.1 & 5771 & 5.0 & 6782 & 5.9\\
 & GodExcludesSpirit & 228 & 0.2 & 274 & 0.2 & 593 & 0.5 & 1137 & 1.0 & 0 & 0.0 & 523 & 0.5 & 1094 & 0.9 & 950 & 0.8 & 666 & 0.6\\
\bottomrule
\end{tabular}
}
\caption{Sample statistics by year}
\end{center}
\end{table}

@@

### Belief in a Spirit/Life Force and/or a God  
We assess belief in a spirit or a God by asking two non-affective questions: "Do you believe in a spirit or Life Force" and "Do you believe in a god?" Responses were coded as 1) **God and Spirit Disbelievers**: those who neither believe in a God nor believe in a Spirit or Life Force; (2) **Spirit Believers/God Disbelievers**: those who disbelieve in a God and believe in a Spirit or Life Force; (3) **God Believers/Spirit Disbelievers**: those who believe in a God but do not believe in a Spirit or Life Force; (4) **God and Spirit Believers**: those who believe in a both God and a Spirit or Life Force.

## Statistical Analyses
Statistical analysis was performed using R version 4.0.2 R version 3.6.1 (2020-06-22) 2019-07-05). We analyze data from participants who responded to the NZAVS survey at least three times between 2010 (Time 2) and 2018 (Time 10), resulting in a sample of  21,705 (*N* = 20,979 with complete responses).  To model growth in subjective well-being we use generalized linear mixed models. In our models, we account for the non-linear effects of time within participants by including the interaction of years (a continuous variable) with Beliefs (a four-level categorical indicator). The baseline for this indicator is disbelief in a Spirit/Life Force and disbelief in a God. To handle the dependencies introduced from the repeated measures, we include individual ID as an effect modeled as random, and we estimated a random effect for the slope for years. This allowed us to model correllation between individual-level intercept (starting high or low in Personal Well-Being/Life Satisfaction) and stability/growth/decline in the slope effect of years. We estimated all statitiscal models using both REML estimation [@R-lme4] and Bayesian estimation [@BRMSpackage] with weakly informative priors. We include the full reports in the Supplement. The REML estimates and the Bayesian estimates converged to within three decimal places.  We include model equations and priors at the end of this article. Here we report the results of the Bayesian estimator.

@@

This method reveals two adjustments sets equired to avoid potential confounding: \{age, employment, male, political conservativism, and time\}, and \{age, ethnicty, male, and time \}.Here, we report the second set because employment is likely to contain more measurement error: some are not employed because they cannot find employment, others because they are not seeking employment (e.g. retirement).  Results were insensitive to the choice set. Indeed, the inference proved robust to the simple model with only the interaction of years and beliefs, and robust to the fully saturated model in which all covariates were included (see the online supplement for details). In the fully saturated model, it appeared that those disbelieving in a Spirit or Life Force, exhibited no growth in Persona Well-Being and Life-Satisifaction over the nine years. Although this result offers stronger stronger evidence for a Spirit/Life Force well-being mechanism, it is likely the effect of collider confounding and masking. We discuss this point in more detail in the supplement, and mention it here to caution researchers from "everything-but-the-kitchen-sink" approaches to regression modelling.

@@

## System and packages
The analysis was performed using R version 4.0.2 (2020-06-22). The Platform was x86_64-apple-darwin17.0 (64-bit) Running under: macOS Catalina 10.15.6
We are greatful to the contributors and mantainers of the following packages:
R [Version 4.0.3; @R-base] and the R-packages *Amelia* [Version 1.7.6; @R-Amelia], *citr* [Version 0.3.2; @R-citr], *dplyr* [Version 1.0.2; @R-dplyr], *equatiomatic* [Version 0.1.0.9000; @R-equatiomatic], *forcats* [Version 0.5.0; @R-forcats], *ggeffects* [Version 1.0.0; @R-ggeffects], *gghighlight* [Version 0.3.0; @R-gghighlight], *ggplot2* [Version 3.3.2; @R-ggplot2], *ggsci* [Version 2.9; @R-ggsci], *lme4* [Version 1.1.25; @R-lme4], *Matrix* [Version 1.2.18; @R-Matrix], *papaja* [Version 0.1.0.9997; @R-papaja], *parameters* [Version 0.9.0.1; @R-parameters], *patchwork* [Version 1.1.0; @R-patchwork], *prettycode* [@R-prettycode], *purrr* [Version 0.3.4; @R-purrr], *Rcpp* [Version 1.0.5; @R-Rcpp_a; @R-Rcpp_b], *readr* [Version 1.4.0; @R-readr], *report* [Version 0.2.0; @R-report], *see* [Version 0.6.0.1; @R-see], *sjPlot* [Version 2.8.6; @R-sjPlot], *stringr* [Version 1.4.0; @R-stringr], *styler* [Version 1.3.2; @R-styler], *table1* [Version 1.2.1; @R-table1], *texreg* [Version 1.37.5; @R-texreg], *tibble* [Version 3.0.4; @R-tibble], *tidyr* [Version 1.1.2; @R-tidyr], and *tidyverse* [Version 1.3.0; @R-tidyverse]

@@

\begin{table}
\begin{center}
\scalebox{0.6}{
\begin{tabular}{l c c}
\toprule
 & PWI & Life Sat \\
\midrule
Intercept                           & $\mathbf{7.14}$   & $\mathbf{5.29}$   \\
                                    & $ [ 7.11;  7.18]$ & $ [ 5.26;  5.31]$ \\
yearsC                              & $\mathbf{0.02}$   & $\mathbf{0.01}$   \\
                                    & $ [ 0.01;  0.02]$ & $ [ 0.00;  0.01]$ \\
Beliefs\_SpiritExcludesGod\_        & $\mathbf{-0.06}$  & $0.01$            \\
                                    & $ [-0.09; -0.03]$ & $ [-0.01;  0.03]$ \\
BeliefsGodAndSpirit                 & $\mathbf{-0.06}$  & $\mathbf{0.03}$   \\
                                    & $ [-0.09; -0.03]$ & $ [ 0.01;  0.06]$ \\
BeliefsGodExcludesSpirit            & $-0.00$           & $\mathbf{0.06}$   \\
                                    & $ [-0.04;  0.04]$ & $ [ 0.03;  0.09]$ \\
Age\_in\_Decades\_C                 & $\mathbf{0.07}$   & $\mathbf{0.08}$   \\
                                    & $ [ 0.06;  0.09]$ & $ [ 0.07;  0.09]$ \\
Ethnic\_CategoriesMaori             & $\mathbf{-0.38}$  & $\mathbf{-0.11}$  \\
                                    & $ [-0.43; -0.33]$ & $ [-0.14; -0.08]$ \\
Ethnic\_CategoriesPacific           & $\mathbf{-0.38}$  & $\mathbf{-0.15}$  \\
                                    & $ [-0.48; -0.29]$ & $ [-0.22; -0.09]$ \\
Ethnic\_CategoriesAsian             & $\mathbf{-0.09}$  & $-0.05$           \\
                                    & $ [-0.18; -0.01]$ & $ [-0.11;  0.02]$ \\
Male1                               & $\mathbf{-0.06}$  & $\mathbf{-0.16}$  \\
                                    & $ [-0.10; -0.02]$ & $ [-0.19; -0.13]$ \\
yearsC:Beliefs\_SpiritExcludesGod\_ & $\mathbf{0.01}$   & $\mathbf{0.01}$   \\
                                    & $ [ 0.00;  0.02]$ & $ [ 0.00;  0.02]$ \\
yearsC:BeliefsGodAndSpirit          & $\mathbf{0.01}$   & $\mathbf{0.01}$   \\
                                    & $ [ 0.01;  0.02]$ & $ [ 0.00;  0.02]$ \\
yearsC:BeliefsGodExcludesSpirit     & $0.00$            & $-0.00$           \\
                                    & $ [-0.01;  0.02]$ & $ [-0.01;  0.01]$ \\
\midrule
SD: Idsd(Intercept)                 & $1.42$            & $0.98$            \\
SD: Idsd(yearsC)                    & $0.13$            & $0.10$            \\
SD: Idcor(Intercept,yearsC)         & $-0.03$           & $0.03$            \\
\bottomrule
\end{tabular}
}
\caption{}
\label{tab:REGRESS_LS}
\end{center}
\end{table}

Figure 1 about here
```{r fig.cap = "Coefficient Graphs", echo =FALSE,out.width="\\textwidth"}
knitr::include_graphics("/Users/jbul716/The\ Virtues\ Project\ Dropbox/Joseph\ Bulbulia/000GIT_STUFF/nzavs_spiritbeliefWellbeing/coefBayesDag.png")
```

@@

```{r fig.cap = "Personal Well-being prediction graph", echo =FALSE, out.width="\\textwidth"}
knitr::include_graphics("/Users/jbul716/The\ Virtues\ Project\ Dropbox/Joseph\ Bulbulia/000GIT_STUFF/nzavs_spiritbeliefWellbeing/pwi_confounder_adjusted.png")
```

>>>>

keywords          : "Belief, God, Health, Longitudinal, Panel, Religion, Spirit, Spirituality, Well-being"
wordcount         : "5181 words in text body, 1191 words in reference section"
bibliography      : ["b-references.bib"]
floatsintext      : no
figurelist        : no
figsintext        : yes
tablelist         : no
footnotelist      : no
linenumbers       : no
mask              : no
draft             : no
documentclass     : "apa6"
classoption       : "man"
output            : papaja::apa6_pdf
extra_dependencies: ["rotating", "booktabs", "graphicx","longtable","amsmath","amsfonts","amssymb"]

---
```{r libraries}
library("tidyverse")
library("papaja")
library("here")
```
```{r global_options, echo=FALSE, include = FALSE}
knitr::opts_chunk$set(#fig.path="Figs/", 
                      message=FALSE, 
                      warning=FALSE,
                      collapse =TRUE,
                      echo=TRUE, 
                      #results="hide", 
                     # fig.width= 10,
                     # fig.height=8,
                      tidy = "styler")
```

@@

Here, we addresses these three challenges of previous research in the following ways: First, to avoid a tautological elision of "spirituality" and "well-being" we adopt a non-affective measure of spirituality as a "believe in a Spirit or Life Force." On the other side, to measure subjective well-being, we use a well-validated scale for personal well-being and a well-validated scale for Life-Satisfaction (see Method). Second, to disentangle spiritual beliefs from traditional religious beliefs and skeptical beliefs we adopt a four-level categorical indicator combining the binary indicator of belief in a Spirit or Life Force with the binary indicator "Do you believe in some form of a spirit or Life Force?" The levels of this four-level categorical indicator for spiritual, religious, and skeptical beliefs are as follows: (1) Skeptics: those who neither believe in a God nor believe in a Spirit or Life Force; (2) Skeptics about God who nevertheless believe in a Spirit or Life Force; (3) Those who believe in a God but do not believe in a Spirit or Life Force; (4) Those who who believe in a both God and a Spirit or Life Force. Clearly such an approach cannot simultaneously address all interests in previous spirituality/well-being research. For example we do not combine our cognitive measure of belief with behavioral measures or attitudinal measures. However, this cost to scope is paid for with benefits to precision. The spiritual indicators measure cognitive states in which distinct varieties of skeptical beliefs, traditional religious beliefs, and non-traditional spirit beliefs are straightforwardly disentangled, and measured within people. Third, to understand how spirit beliefs affect growth, stability, or decline in psychological well-being over time, we leverage longitudinal responses from a nationally diverse probability sample year over nine years (2010 to 2018 New Zealand Attitudes and Values Study, NZAVS).

@@

\begin{table}[ ht ] 
\centering 
\caption{}\label{}
\scalebox{0.6}{
\begin{tabular}{ l c c c c c c c c c }
\toprule
 &   \multicolumn{ 9 }{c}{ Wave }\\ 
  & 2010 & 2011 & 2012 & 2013 & 2014 & 2015 & 2016 & 2017 & 2018 \\ 
 & n = \emph{missing} & n = \emph{missing} & n = 8547 & n = 13472 & n = \emph{missing} & n = 11687 & n = 16048 & n = 14174 & n = 14883 \\ 
 \midrule
Age &   &   &   &   &   &   &   &   &  \\ 
NZdep &   &   &   &   &   &   &   &   &  \\ 
Education &   &   &   &   &   &   &   &   &  \\ 
Employed &   &   &   &   &   &   &   &   &  \\ 
\hspace{6pt}    0 &  (\%) &  (\%) & 2094 (24.5\%) & 2955 (21.9\%) &  (\%) & 2525 (21.6\%) & 3486 (21.7\%) & 3223 (22.7\%) & 3404 (22.9\%)\\ 
\hspace{6pt}    1 &  (\%) &  (\%) & 6453 (75.5\%) & 10517 (78.1\%) &  (\%) & 9162 (78.4\%) & 12562 (78.3\%) & 10951 (77.3\%) & 11479 (77.1\%)\\ 
Ethnic\_Categories &   &   &   &   &   &   &   &   &  \\ 
\hspace{6pt}    Euro &  (\%) &  (\%) & 6731 (78.8\%) & 11024 (81.8\%) &  (\%) & 9671 (82.8\%) & 13389 (83.4\%) & 11814 (83.3\%) & 12427 (83.5\%)\\ 
\hspace{6pt}    Maori &  (\%) &  (\%) & 1188 (13.9\%) & 1608 (11.9\%) &  (\%) & 1341 (11.5\%) & 1749 (10.9\%) & 1620 (11.4\%) & 1677 (11.3\%)\\ 
\hspace{6pt}    Pacific &  (\%) &  (\%) & 284 (3.3\%) & 339 (2.5\%) &  (\%) & 277 (2.4\%) & 331 (2.1\%) & 243 (1.7\%) & 261 (1.8\%)\\ 
\hspace{6pt}    Asian &  (\%) &  (\%) & 344 (4\%) & 501 (3.7\%) &  (\%) & 398 (3.4\%) & 579 (3.6\%) & 497 (3.5\%) & 518 (3.5\%)\\ 
Male &   &   &   &   &   &   &   &   &  \\ 
\hspace{6pt}    0 &  (\%) &  (\%) & 5302 (62\%) & 8457 (62.8\%) &  (\%) & 7325 (62.7\%) & 10148 (63.2\%) & 9011 (63.6\%) & 9433 (63.4\%)\\ 
\hspace{6pt}    1 &  (\%) &  (\%) & 3245 (38\%) & 5015 (37.2\%) &  (\%) & 4362 (37.3\%) & 5900 (36.8\%) & 5163 (36.4\%) & 5450 (36.6\%)\\ 
Has\_Partner &   &   &   &   &   &   &   &   &  \\ 
\hspace{6pt}    0 &  (\%) &  (\%) & 2408 (28.2\%) & 3546 (26.3\%) &  (\%) & 2815 (24.1\%) & 3773 (23.5\%) & 3330 (23.5\%) & 3556 (23.9\%)\\ 
\hspace{6pt}    1 &  (\%) &  (\%) & 6139 (71.8\%) & 9926 (73.7\%) &  (\%) & 8872 (75.9\%) & 12275 (76.5\%) & 10844 (76.5\%) & 11327 (76.1\%)\\ 
Political\_Conservativism &   &   &   &   &   &   &   &   &  \\ 
Urban &   &   &   &   &   &   &   &   &  \\ 
\hspace{6pt}    Not\_Urban &  (\%) &  (\%) & 2865 (33.5\%) & 4278 (31.8\%) &  (\%) & 3951 (33.8\%) & 5538 (34.5\%) & 2610 (18.4\%) & 2661 (17.9\%)\\ 
\hspace{6pt}    Urban &  (\%) &  (\%) & 5682 (66.5\%) & 9194 (68.2\%) &  (\%) & 7736 (66.2\%) & 10510 (65.5\%) & 11564 (81.6\%) & 12222 (82.1\%)\\ 
Beliefs &   &   &   &   &   &   &   &   &  \\ 
\hspace{6pt}    \_Skeptic\_ &  (\%) &  (\%) & 2048 (24\%) & 3286 (24.4\%) &  (\%) & 2951 (25.3\%) & 4116 (25.6\%) & 3916 (27.6\%) & 4120 (27.7\%)\\ 
\hspace{6pt}    \_SpiritExcludesGod\_ &  (\%) &  (\%) & 2295 (26.9\%) & 3735 (27.7\%) &  (\%) & 3169 (27.1\%) & 4501 (28\%) & 3918 (27.6\%) & 3970 (26.7\%)\\ 
\hspace{6pt}    GodAndSpirit &  (\%) &  (\%) & 3707 (43.4\%) & 5499 (40.8\%) &  (\%) & 5119 (43.8\%) & 6453 (40.2\%) & 5464 (38.5\%) & 6210 (41.7\%)\\ 
\hspace{6pt}    GodExcludesSpirit &  (\%) &  (\%) & 497 (5.8\%) & 952 (7.1\%) &  (\%) & 448 (3.8\%) & 978 (6.1\%) & 876 (6.2\%) & 583 (3.9\%)\\ 
LIFESAT &   &   &   &   &   &   &   &   &  \\ 
PWI &   &   &   &   &   &   &   &   &  \\ 
\bottomrule

\end{tabular}
}
\end{table}

@@

### Belief in a Spirit/Life Force and/or a God  
We assess belief in a spirit or a God by asking two non-affective questions: "Do you believe in a spirit or Life Force" and "Do you believe in a god?" Responses were coded as (1) Skeptics: those who neither believe in a God nor believe in a Spirit or Life Force; (2) Skeptics about God who believe in a Spirit or Life Force; (3) Believers in a God who disbelieve in a Spirit or Life Force; (4) Believers in a God and Spirit or Life Force. These believe were developed for the NZAVS from the 2005 Eurobarometer.

## Statistical Analyses
Statistical analysis was performed using R version 4.0.2 R version 3.6.1 (2020-06-22) 2019-07-05). We analyze data from participants who responded to the NZAVS survey at least three times between 2010 (Time 2) and 2018 (Time 10), resulting in a sample of  21,705 (*N* = 20,979 with complete responses).  To model growth in subjective well-being we use generalized linear mixed models. In our models, we account for the non-linear effects of time within participants by including the interaction of years (a continuous variable) with Beliefs (a four-level categorical indicator). The baseline for this indicator is Skepticism about a Spirit/Life Forces and Skepticism about a God. To handle the dependencies introduced from the repeated measures, we include individual ID as an effect modeled as random, and we estimated a random effect for the slope for years. This allowed us to model correllation between individual-level intercept (starting high or low in Personal Well-Being/Life Satisfaction) and stability/growth/decline in the slope effect of years. We estimated all statitiscal models using both REML estimation [@R-lme4] and Bayesian estimation [@BRMSpackage] with weakly informative priors. We include the full reports in the Supplement. The REML estimates and the Bayesian estimates converged to within three decimal places.  We include model equations and priors at the end of this article. Here we report the results of the Bayesian estimator.

@@

This method reveals two adjustments sets equired to avoid potential confounding: \{age, employment, male, political conservativism, and time\}, and \{age, ethnicty, male, and time \}.Here, we report the second set because employment is likely to contain more measurement error: some are not employed because they cannot find employment, others because they are not seeking employment (e.g. retirement).  Results were insensitive to the choice set. Indeed, the inference proved robust to the simple model with only the interaction of years and beliefs, and robust to the fully saturated model in which all covariates were included (see the online supplement for details). In the fully saturated model, it appeared that those skeptical of a Spirit or Life Force, exhibited no growth in Persona Well-Being and Life-Satisifaction over the nine years. Although this result offers stronger stronger evidence for a Spirit/Life Force well-being mechanism, it is likely the effect of collider confounding and masking. We discuss this point in more detail in the supplement, and mention it here to caution researchers from "everything-but-the-kitchen-sink" approaches to regression modelling.

@@

## System and packages
The analysis was performed using R version 4.0.2 (2020-06-22). The Platform was x86_64-apple-darwin17.0 (64-bit) Running under: macOS Catalina 10.15.6
We are greatful to the contributors and mantainers of the following packages:
R [Version 4.0.2; @R-base] and the R-packages *Amelia* [Version 1.7.6; @R-Amelia], *citr* [Version 0.3.2; @R-citr], *dplyr* [Version 1.0.2; @R-dplyr], *equatiomatic* [Version 0.1.0.9000; @R-equatiomatic], *forcats* [Version 0.5.0; @R-forcats], *ggeffects* [Version 0.15.0; @R-ggeffects], *gghighlight* [Version 0.3.0; @R-gghighlight], *ggplot2* [Version 3.3.2; @R-ggplot2], *ggsci* [Version 2.9; @R-ggsci], *lme4* [Version 1.1.23; @R-lme4], *Matrix* [Version 1.2.18; @R-Matrix], *papaja* [Version 0.1.0.9997; @R-papaja], *parameters* [Version 0.8.6.1; @R-parameters], *patchwork* [Version 1.0.1; @R-patchwork], *prettycode* [Version 1.1.0; @R-prettycode], *purrr* [Version 0.3.4; @R-purrr], *Rcpp* [Version 1.0.5.1; @R-Rcpp_a; @R-Rcpp_b], *readr* [Version 1.3.1; @R-readr], *report* [Version 0.1.0; @R-report], *see* [Version 0.5.2.1; @R-see], *sjPlot* [Version 2.8.4; @R-sjPlot], *stringr* [Version 1.4.0; @R-stringr], *styler* [Version 1.3.2; @R-styler], *table1* [Version 1.2; @R-table1], *texreg* [Version 1.37.5; @R-texreg], *tibble* [Version 3.0.4; @R-tibble], *tidyr* [Version 1.1.2; @R-tidyr], and *tidyverse* [Version 1.3.0; @R-tidyverse]"

@@

\begin{table}
\begin{center}
\scalebox{0.5}{
\begin{tabular}{l c c}
\toprule
 & PWI & Life Sat \\
\midrule
Intercept                           & $\mathbf{7.14}$   & $\mathbf{5.29}$   \\
                                    & $ [ 7.11;  7.18]$ & $ [ 5.26;  5.31]$ \\
yearsC                              & $\mathbf{0.02}$   & $\mathbf{0.01}$   \\
                                    & $ [ 0.01;  0.02]$ & $ [ 0.00;  0.01]$ \\
Beliefs\_SpiritExcludesGod\_        & $\mathbf{-0.06}$  & $0.01$            \\
                                    & $ [-0.09; -0.03]$ & $ [-0.01;  0.03]$ \\
BeliefsGodAndSpirit                 & $\mathbf{-0.06}$  & $\mathbf{0.03}$   \\
                                    & $ [-0.09; -0.03]$ & $ [ 0.01;  0.06]$ \\
BeliefsGodExcludesSpirit            & $-0.00$           & $\mathbf{0.06}$   \\
                                    & $ [-0.04;  0.04]$ & $ [ 0.03;  0.09]$ \\
Age\_in\_Decades\_C                 & $\mathbf{0.07}$   & $\mathbf{0.08}$   \\
                                    & $ [ 0.06;  0.09]$ & $ [ 0.07;  0.09]$ \\
Ethnic\_CategoriesMaori             & $\mathbf{-0.38}$  & $\mathbf{-0.11}$  \\
                                    & $ [-0.43; -0.33]$ & $ [-0.14; -0.08]$ \\
Ethnic\_CategoriesPacific           & $\mathbf{-0.38}$  & $\mathbf{-0.15}$  \\
                                    & $ [-0.48; -0.29]$ & $ [-0.22; -0.09]$ \\
Ethnic\_CategoriesAsian             & $\mathbf{-0.09}$  & $-0.05$           \\
                                    & $ [-0.18; -0.01]$ & $ [-0.11;  0.02]$ \\
Male1                               & $\mathbf{-0.06}$  & $\mathbf{-0.16}$  \\
                                    & $ [-0.10; -0.02]$ & $ [-0.19; -0.13]$ \\
yearsC:Beliefs\_SpiritExcludesGod\_ & $\mathbf{0.01}$   & $\mathbf{0.01}$   \\
                                    & $ [ 0.00;  0.02]$ & $ [ 0.00;  0.02]$ \\
yearsC:BeliefsGodAndSpirit          & $\mathbf{0.01}$   & $\mathbf{0.01}$   \\
                                    & $ [ 0.01;  0.02]$ & $ [ 0.00;  0.02]$ \\
yearsC:BeliefsGodExcludesSpirit     & $0.00$            & $-0.00$           \\
                                    & $ [-0.01;  0.02]$ & $ [-0.01;  0.01]$ \\
\midrule
SD: Idsd(Intercept)                 & $1.42$            & $0.98$            \\
SD: Idsd(yearsC)                    & $0.13$            & $0.10$            \\
SD: Idcor(Intercept,yearsC)         & $-0.03$           & $0.03$            \\
\bottomrule
\end{tabular}
}
\caption{}
\label{tab:REGRESS_LS}
\end{center}
\end{table}

Figure 1 about here
```{r fig.cap = "Coefficient Graphs", echo =FALSE,out.width="\\textwidth"}
knitr::include_graphics("/Users/jbul176/The\ Virtues\ Project\ Dropbox/Joseph\ Bulbulia/000GIT_STUFF/nzavs_spiritbeliefWellbeing/coefBayesDag.png")
```

@@

```{r fig.cap = "Personal Well-being prediction graph", echo =FALSE, out.width="\\textwidth"}
knitr::include_graphics("/Users/jbul176/The\ Virtues\ Project\ Dropbox/Joseph\ Bulbulia/000GIT_STUFF/nzavs_spiritbeliefWellbeing/pwi_confounder_adjusted.png")
```

The effect of years $\times$ belief in a *Spirit Excludes God*  is reliably positive  (posterior median = 0.011, HDI [ 0.003,  0.018], pd = 99.02%). We infer that people with beliefs in a spirit or Life Force who do not believe in a God tended to grow in their personal well-being over time at a faster rate than skeptics grew.
  - The effect of years $\times$ in *God And Spirit*  is reliably positive  (posterior median = 0.014, HDI [0.007, 0.021], pd = 99.98%). We infer that people who believe in a spirit or Life Force who also believe in a God tended to grow in their personal well-being over time at a faster rate than skeptics grew.
  - The effect of years $\times$ belief in *God Excludes Spirit*  is not reliable (posterior median = 0.004, HDI [-0.009,  0.015], pd =  68.73%).  We infer that people who believe in a God but not a spirit or Life Force did not grow at a faster rate in personal well-being than skeptics grew.
